---
title: "Analysis of the connection between sleep and menopause"
output:
  html_document:
    df_print: paged
---

<!-- ============================================================================
  02 - Perimenopausal cohort: preprocessing and ordinal regression
  Project: sleepMenopause - androgen decline and perimenopausal sleep
============================================================================ -->

## Overview

Part of the **sleepMenopause** project (androgen decline and perimenopausal
sleep-onset disruption).

This notebook cleans the perimenopausal sleep-survey responses (consent and
completion filtering, sex/gender consistency checks), derives the analysis
dataset, and models associations between hormonal use, clinical triggers and
individual sleep outcomes using ordinal logistic regression (`MASS::polr`) and
Spearman correlations. Corresponds to manuscript **Figures 2-4**.

- **Inputs:** `dt/Sleep Survey_*.csv` (raw survey exports — *not included*; see
  *Data availability* in `README.md`).
- **Outputs:** cleaned analysis tables (`dt_out/`) and regression / correlation
  results.
- **Dependencies:** dplyr, tidyr, caret, ggplot2, stringr, ggbiplot, MASS, broom.

> Raw participant data are not distributed with this repository.

---

```{r}
library(dplyr)
library(tidyr)
library(caret)
library(ggplot2)
library(stringr)
library(ggbiplot)
library(MASS)
library(broom)
```

## Preprocessing 

```{r}
dt = read.csv("dt/Sleep Survey_June 20, 2024_09.13.csv", fileEncoding="latin1")
columns = dt[1,]
dt = dt[-1,]
colnames(dt) <- columns
```
Ensure consent
<br>
Columns linked to consent: c(18:23)]
```{r}
dt_consent = dt
for (consent_col in c(18:23)){
  dt_consent = dt_consent[dt_consent[,consent_col] == "I confirm",]
}
  
```

Ensure completed questionnaire 
```{r}
dt_complete = dt_consent[dt_consent$Progress ==100,]
```

Check sex/gender change 
```{r}
print(sum(dt_complete$`What sex were you assigned at birth?` == dt_complete$`How do you describe yourself?`))
print(nrow(dt_complete))
```

```{r}
dt_gender = dt_complete[dt_complete$`What sex were you assigned at birth?` == dt_complete$`How do you describe yourself?`,]
```

```{r}
write.csv(dt_gender,"dt_out/dt_cleaned_v1.csv")
```

```{r}
dt_clean = dt_gender[,-c(1:25)]
write.csv(dt_clean,"dt_out/dt_cleaned_v2.csv")
```

Feature derivation
```{r}
data = dt_clean

data <- data %>%
  mutate(across(where(is.character), as.factor))  # Convert character columns to factors


```

## Recode all variables
```{r}
sleep_time_mapping <- setNames(c(1, 2, 3, 4, 5),
                               c("Less than 5 minutes", "5-15 minutes", "15-30 minutes", "30-60 minutes", "More than 60 minutes"))

# Convert to numerical ranks using the mapping
data$sleep_time_to_fall_asleep_numeric <- as.numeric(sapply(data$`How long does it take you to fall asleep?`, function(x) sleep_time_mapping[x]))


data$age_numeric <- case_when(
  data$`How old are you? Please select one option` == "18-29" ~ 23.5,
  data$`How old are you? Please select one option` == "30-39" ~ 34.5,
  data$`How old are you? Please select one option` == "40-49" ~ 44.5,
  data$`How old are you? Please select one option` == "50-59" ~ 54.5,
  data$`How old are you? Please select one option` == "60-69" ~ 64.5,
  data$`How old are you? Please select one option` == "70+" ~ 75,  # Assuming 75 as midpoint for 70+
  TRUE ~ NA_real_  # Handles other or missing categories
)
data$white_british = data$`How would you describe your ethnic origin? Please select one option - Selected Choice` == "White English / Welsh / Scottish / Northern Irish / British"
```


```{r}
data$income_numeric <- case_when(
  data$`What is your annual income? (We ask this question to better understand how socioeconomic factors may influence sleep patterns during menopause, to inform recommendations and therapies for improved well-being. Please only answer if you are happy to do so):` == "<£10,000" ~ 8000,   # Assuming midpoint for less than £10,000 is £8,000
  data$`What is your annual income? (We ask this question to better understand how socioeconomic factors may influence sleep patterns during menopause, to inform recommendations and therapies for improved well-being. Please only answer if you are happy to do so):` == "£10,000 - £20,000" ~ 15000,  # Midpoint of £10,000 and £20,000
  data$`What is your annual income? (We ask this question to better understand how socioeconomic factors may influence sleep patterns during menopause, to inform recommendations and therapies for improved well-being. Please only answer if you are happy to do so):` == "£20,000 - £30,000" ~ 25000,  # Midpoint of £20,000 and £30,000
  data$`What is your annual income? (We ask this question to better understand how socioeconomic factors may influence sleep patterns during menopause, to inform recommendations and therapies for improved well-being. Please only answer if you are happy to do so):` == "£30,000 - £40,000" ~ 35000,  # Midpoint of £30,000 and £40,000
  data$`What is your annual income? (We ask this question to better understand how socioeconomic factors may influence sleep patterns during menopause, to inform recommendations and therapies for improved well-being. Please only answer if you are happy to do so):` == "£40,000 - £60,000" ~ 50000,  # Midpoint of £40,000 and £60,000
  data$`What is your annual income? (We ask this question to better understand how socioeconomic factors may influence sleep patterns during menopause, to inform recommendations and therapies for improved well-being. Please only answer if you are happy to do so):` == "£60,000 - £80,000" ~ 70000,  # Midpoint of £60,000 and £80,000
  data$`What is your annual income? (We ask this question to better understand how socioeconomic factors may influence sleep patterns during menopause, to inform recommendations and therapies for improved well-being. Please only answer if you are happy to do so):` == ">£60,000" ~ 80000,   # Assuming a reasonable midpoint for incomes above £60,000
  data$`What is your annual income? (We ask this question to better understand how socioeconomic factors may influence sleep patterns during menopause, to inform recommendations and therapies for improved well-being. Please only answer if you are happy to do so):` == "Prefer not to say" ~ NA_real_,  # Handle 'Prefer not to say' as NA
  TRUE ~ NA_real_  # Handles other or missing categories
)
```

```{r}
data$sleep_quality <- case_when(
  data$`How would you rate your sleep quality?` == "Very poor" ~ 1,
  data$`How would you rate your sleep quality?` == "Poor" ~ 2,
  data$`How would you rate your sleep quality?` == "Average" ~ 3,
  data$`How would you rate your sleep quality?` == "Good" ~ 4,
  data$`How would you rate your sleep quality?` == "Very good" ~ 5,
  TRUE ~ NA_real_  # Handles missing categories or unexpected values
)
```

```{r}
data$time_to_fall_asleep_numeric <- case_when(
  data$`How long does it take you to fall asleep?` == "<15 mins " ~ 1,
  data$`How long does it take you to fall asleep?` == "15-30 mins " ~ 2,
  data$`How long does it take you to fall asleep?` == "30-45 mins " ~ 3,
  data$`How long does it take you to fall asleep?` == "45-60mins " ~ 4,
  data$`How long does it take you to fall asleep?` == ">1 hour" ~ 5,
  TRUE ~ NA_real_  # Handles missing categories or unexpected values
)
```


```{r}
data$awake_minutes_numeric <- case_when(
  data$`If you wake in the night, how long in total are you awake?` == "<15 mins" ~ 7.5,
  data$`If you wake in the night, how long in total are you awake?` == "<30 mins" ~ 22.5,
  data$`If you wake in the night, how long in total are you awake?` == "30-60 mins" ~ 45,
  data$`If you wake in the night, how long in total are you awake?` == "1-2 hours" ~ 90,
  data$`If you wake in the night, how long in total are you awake?` == "2-3 hours" ~ 150,
  data$`If you wake in the night, how long in total are you awake?` == ">3 hours" ~ 180,  # Could be more, but 180 is a reasonable estimate for over 3 hours
  data$`If you wake in the night, how long in total are you awake?` == "I don't wake in the night" ~ 0,
  TRUE ~ NA_real_  # Handles missing categories or unexpected values
)

```

```{r}
choices <- c("No reason you can think of, you just wake up",
             "Physical symptoms, e.g, pain, hot flushes, sweats, racing heart, needing to use the bathroom",
             "Having worries or problems playing on your mind",
             "The person I share a bed with, e.g. due to their snoring or movement etc.")

# Generate binary variables for each choice
for(choice in choices) {
  data[[str_replace_all(choice, "[^A-Za-z0-9]", "_")]] <- as.integer(str_detect(data$`What wakes you during the night? (Tick all that apply) - Selected Choice`, fixed(choice)))
}

```

```{r}
data$early_waking_numeric <- case_when(
  data$`Do you have a problem with waking too early and not getting back to sleep?` == "Not at all  " ~ 0,
  data$`Do you have a problem with waking too early and not getting back to sleep?` == "Sometimes  " ~ 1,
  data$`Do you have a problem with waking too early and not getting back to sleep?` == "Frequently  " ~ 2,
  TRUE ~ NA_real_  # Handles missing categories or unexpected values
)
```

```{r}
data$functional_impairment_numeric <- case_when(
  data$`To what extent does your sleep in interfere with daily functioning, e.g., day-time tiredness, mood, memory, work, concentration` == "Not at all " ~ 0,
  data$`To what extent does your sleep in interfere with daily functioning, e.g., day-time tiredness, mood, memory, work, concentration` == "A little " ~ 1,
  data$`To what extent does your sleep in interfere with daily functioning, e.g., day-time tiredness, mood, memory, work, concentration` == "Somewhat " ~ 2,
  data$`To what extent does your sleep in interfere with daily functioning, e.g., day-time tiredness, mood, memory, work, concentration` == "Interferes a lot " ~ 3,
  TRUE ~ NA_real_  # Handles missing categories or unexpected values
)

# Check the converted data
head(data[, c("To what extent does your sleep in interfere with daily functioning, e.g., day-time tiredness, mood, memory, work, concentration", "functional_impairment_numeric")])

```


```{r}
data$daytime_sleepiness_numeric <- case_when(
  data$`Do you "nod off", doze or fall asleep during the day without obvious reason?` == "Never" ~ 0,
  data$`Do you "nod off", doze or fall asleep during the day without obvious reason?` == "Sometimes" ~ 1,
  data$`Do you "nod off", doze or fall asleep during the day without obvious reason?` == "Regularly" ~ 2,
  data$`Do you "nod off", doze or fall asleep during the day without obvious reason?` == "Always" ~ 3,
  TRUE ~ NA_real_  # Handles missing categories or unexpected values
)

# Check the converted data
head(data[, c("Do you \"nod off\", doze or fall asleep during the day without obvious reason?", "daytime_sleepiness_numeric")])

```

```{r}
data$exercise_frequency_numeric <- case_when(
  data$`How often do you exercise or engage in physical activity that raises the heart rate for at least 30 minutes:` == "None at all" ~ 0,
  data$`How often do you exercise or engage in physical activity that raises the heart rate for at least 30 minutes:` == "Once a week" ~ 1,
  data$`How often do you exercise or engage in physical activity that raises the heart rate for at least 30 minutes:` == "2-3 times per week" ~ 2,
  data$`How often do you exercise or engage in physical activity that raises the heart rate for at least 30 minutes:` == "4-5 times per week" ~ 3,
  data$`How often do you exercise or engage in physical activity that raises the heart rate for at least 30 minutes:` == "Most days" ~ 4,
  TRUE ~ NA_real_  # Handles missing categories or unexpected values
)

# Check the converted data
head(data[, c("How often do you exercise or engage in physical activity that raises the heart rate for at least 30 minutes:", "exercise_frequency_numeric")])

```

```{r}
data$take_HRT = data$`Do you take HRT? (Tick all that apply) - Selected Choice` != "I do not take HRT"
```

```{r}
data$take_testosterone = data$`Do you take testosterone?` =="Yes "

```

```{r}
choices <- c(
  "Combined HRT (oestrogen and progesterone) in tablet form, e.g., Kliovance, Elleste Duet, Bijuve, Femoston",
  "Combined HRT (oestrogen and progesterone) patch e.g., Evorel Sequi, Evorel Conti",
  "Oestrogen only gel or patch",
  "Vaginal progesterone e.g. Utrogestan, Cyclogest, Lutigest",
  "Other",
  "I do not take HRT",
  "I used to take HRT but do not anymore",
  "Utrogestan or Gepretix (taken orally)"
)

# Generate binary variables for each choice
for(choice in choices) {
  data[[str_replace_all(choice, "[^A-Za-z0-9]", "_")]] <- as.integer(str_detect(data$`Do you take HRT? (Tick all that apply) - Selected Choice`, fixed(choice)))
}

# Check the new columns
head(data[, c("Do you take HRT? (Tick all that apply) - Selected Choice", grep("HRT", names(data), value = TRUE))])

```

```{r}
data$bmi = as.numeric(as.character(data$`Do you know your BMI (body mass index)? (We ask this question as body mass can impact on our sleep. Please only answer if you are happy to do so):`))
```

```{r}
write.csv(data, "dt_out/dt_cleaned_v3.csv")
```

### Remove participants below 49 years old 

```{r}
data = subset(data, age_numeric > 40)
write.csv(data, "dt_out/dt_cleaned_v3_over40.csv")
```

## Analysis

```{r}
pca_data <- subset(data, select=c(age_numeric, sleep_time_to_fall_asleep_numeric, income_numeric, awake_minutes_numeric,
                   early_waking_numeric, functional_impairment_numeric, daytime_sleepiness_numeric, exercise_frequency_numeric))

# Check for NA values and handle them
pca_data <- na.omit(pca_data)

# Standardize the data
pca_data_scaled <- scale(pca_data)

# Perform PCA
pca_result <- prcomp(pca_data_scaled, center = TRUE, scale. = TRUE)

# Plot PCA
plot_ggbiplot <- ggbiplot(pca_result, obs.scale = 1, var.scale = 1, ellipse = TRUE, circle = TRUE) # replace 'group_variable' with actual group variable if needed
print(plot_ggbiplot)
```

```{r}
loadings <- pca_result$rotation[, 1]  # Only for the first principal component

# Convert loadings to a data frame for easier handling
loadings_df <- as.data.frame(loadings)
loadings_df$variable <- rownames(loadings_df)
colnames(loadings_df) <- c("loading", "variable")

# Ordering loadings by absolute value for better visualization
loadings_df <- loadings_df %>%
  arrange(desc(abs(loading)))
```

```{r}
ggplot(loadings_df, aes(x = reorder(variable, loading), y = loading, fill = loading > 0)) +
  geom_col() +
  coord_flip() +  # Flips the axes for better readability of variable names
  labs(title = "Contributions of Variables to PC1",
       x = "Variable",
       y = "Loading") +
  scale_fill_manual(name = "Direction", 
                    labels = c("Negative", "Positive"), 
                    values = c("red", "blue")) +  # Colors to distinguish positive and negative contributions
  theme_minimal()
```


```{r}
pca_data <- subset(data, select=c(sleep_time_to_fall_asleep_numeric, awake_minutes_numeric,
                   early_waking_numeric, functional_impairment_numeric, daytime_sleepiness_numeric, exercise_frequency_numeric))

pca_data$take_testosterone <- data$take_testosterone
pca_data <- na.omit(pca_data)

# Standardize the data for PCA
pca_data_scaled <- scale(subset(pca_data, select=c(-take_testosterone)))
pca_data <- na.omit(pca_data)

# Perform PCA
pca_result <- prcomp(pca_data_scaled, center = TRUE, scale. = TRUE)

# Plot PCA with ggbiplot and color by HRT and testosterone status
plot_ggbiplot <- ggbiplot(pca_result, obs.scale = 1, var.scale = 1,
                          groups = pca_data$take_testosterone, ellipse = TRUE, circle = TRUE)
print(plot_ggbiplot)

```

```{r}
pca_data <- subset(data,select =c(sleep_time_to_fall_asleep_numeric, awake_minutes_numeric,
                   early_waking_numeric, functional_impairment_numeric, daytime_sleepiness_numeric, exercise_frequency_numeric))

pca_data$take_HRT <- data$take_HRT
pca_data <- na.omit(pca_data)

# Standardize the data for PCA
pca_data_scaled <- scale(subset(pca_data, select =c(-take_HRT)))
pca_data <- na.omit(pca_data)

# Perform PCA
pca_result <- prcomp(pca_data_scaled, center = TRUE, scale. = TRUE)

# Plot PCA with ggbiplot and color by HRT and testosterone status
plot_ggbiplot <- ggbiplot(pca_result, obs.scale = 1, var.scale = 1,
                          groups = pca_data$take_HRT, ellipse = TRUE, circle = TRUE)
print(plot_ggbiplot)
```
```{r}
pca_data <- subset(data, select =c(sleep_time_to_fall_asleep_numeric, awake_minutes_numeric,
                   early_waking_numeric, functional_impairment_numeric, daytime_sleepiness_numeric, exercise_frequency_numeric))

pca_data$Physical_symptoms__e_g__pain__hot_flushes__sweats__racing_heart__needing_to_use_the_bathroom <- data$Physical_symptoms__e_g__pain__hot_flushes__sweats__racing_heart__needing_to_use_the_bathroom
pca_data <- na.omit(pca_data)

# Standardize the data for PCA
pca_data_scaled <- scale(subset(pca_data, select=c(-Physical_symptoms__e_g__pain__hot_flushes__sweats__racing_heart__needing_to_use_the_bathroom)))
pca_data <- na.omit(pca_data)

# Perform PCA
pca_result <- prcomp(pca_data_scaled, center = TRUE, scale. = TRUE)

# Plot PCA with ggbiplot and color by HRT and testosterone status
plot_ggbiplot <- ggbiplot(pca_result, obs.scale = 1, var.scale = 1,
                          groups = pca_data$Physical_symptoms__e_g__pain__hot_flushes__sweats__racing_heart__needing_to_use_the_bathroom, ellipse = TRUE, circle = TRUE)
print(plot_ggbiplot)
```

```{r}

umap_result <- umap::umap(as.matrix(pca_data_scaled))
umap_df <- as.data.frame(umap_result$layout)
colnames(umap_df) <- c("UMAP1", "UMAP2")

# Optionally, if you have a categorical variable to color by (e.g., clusters or groups):
umap_df$group <- data$take_testosterone  # make sure this aligns by row with your umap data

# Plot
ggplot(umap_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = factor(umap_df$group)), alpha = 0.5) +  # Remove color mapping if not needed
  labs(title = "UMAP Projection", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal() +
  scale_color_viridis_d()  # Only if you have groups to color
```

```{r}
data_pca = data
data_pca$pc1 = pca_result$x[,1]

model <- lm(pc1 ~ age_numeric + income_numeric + exercise_frequency_numeric + bmi + take_HRT, data = data_pca)
summary(model)
  
```
```{r}
model <- lm(pc1 ~ age_numeric + income_numeric+ + exercise_frequency_numeric+ bmi+ take_testosterone, data = data_pca)
summary(model)
```
```{r}
model <- lm(pc1 ~ age_numeric + income_numeric + exercise_frequency_numeric+ bmi + Physical_symptoms__e_g__pain__hot_flushes__sweats__racing_heart__needing_to_use_the_bathroom, data = data_pca)
summary(model)
```

## Models

```{r}
# Define the list of response variables
response_vars <- c("sleep_quality", "time_to_fall_asleep_numeric", "awake_minutes_numeric", 
                   "early_waking_numeric", "functional_impairment_numeric", "daytime_sleepiness_numeric")

# Initialize an empty data frame to store coefficients, confidence intervals, and p-values
coefficients_df <- data.frame(variable = character(), coefficient = numeric(), 
                              ci_lower = numeric(), ci_upper = numeric(), p_value = numeric())

# Loop through each response variable, fit the model, and extract the coefficient, CI, and p-value for 'take_HRT'
for (var in response_vars) {
  data[,var] = as.numeric(as.character(data[,var]))
  data[[var]] <- (data[[var]] - min(data[[var]], na.rm = TRUE)) / (max(data[[var]], na.rm = TRUE) - min(data[[var]], na.rm = TRUE))
  formula <- as.formula(paste(var, "~ age_numeric + income_numeric + exercise_frequency_numeric + take_HRT"))
  model <- lm(formula, data = data)
  
  # Extracting the coefficient, confidence interval, and p-value for 'take_HRT'
  summary_model <- summary(model)
  take_HRT_coef <- coef(summary_model)["take_HRTTRUE", ]
  ci <- confint.default(model, "take_HRTTRUE", level = 0.95)

  # Append the results
  coefficients_df <- rbind(coefficients_df, data.frame(variable = var, 
                                                       coefficient = take_HRT_coef["Estimate"],
                                                       ci_lower = ci[1],
                                                       ci_upper = ci[2],
                                                       p_value = take_HRT_coef["Pr(>|t|)"]))
}
# Define significance levels for coloring
coefficients_df$significance <- ifelse(coefficients_df$p_value < 0.001, "p < 0.001",
                               ifelse(coefficients_df$p_value < 0.01, "p < 0.01",
                               ifelse(coefficients_df$p_value < 0.05, "p < 0.05", "p >= 0.05")))

# Plotting the coefficients as a forest plot
ggplot(coefficients_df, aes(x = reorder(variable,-coefficient), y = coefficient, color = significance)) +
  geom_point(size = 3) +  # Display the coefficient estimates as points
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +  # Display error bars for CIs
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_color_manual(values = c("p < 0.001" = "red", "p < 0.01" = "orange", "p < 0.05" = "blue", "p >= 0.05" = "grey")) +
  labs(title = "Effect of HRT on Various Sleep and Activity Measures",
       subtitle = "Colored by Significance Levels of P-Value",
       x = "Coefficient of 'take_HRT'",
       y = "Response Variable") +
  coord_flip()+
  theme_minimal()
```

```{r}
# Define the list of response variables
response_vars <- c("sleep_quality", "time_to_fall_asleep_numeric", "awake_minutes_numeric", 
                   "early_waking_numeric", "functional_impairment_numeric", "daytime_sleepiness_numeric")

# Initialize an empty data frame to store coefficients, confidence intervals, and p-values
coefficients_df <- data.frame(variable = character(), coefficient = numeric(), 
                              ci_lower = numeric(), ci_upper = numeric(), p_value = numeric())

# Loop through each response variable, fit the model, and extract the coefficient, CI, and p-value for 'take_testosterone'
for (var in response_vars) {
  data[[var]] <- (data[[var]] - min(data[[var]], na.rm = TRUE)) / (max(data[[var]], na.rm = TRUE) - min(data[[var]], na.rm = TRUE))
  formula <- as.formula(paste(var, "~ age_numeric + income_numeric + white_british + exercise_frequency_numeric + take_testosterone"))
  model <- lm(formula, data = data)
  
  # Extracting the coefficient, confidence interval, and p-value for 'take_testosterone'
  summary_model <- summary(model)
  coef <- coef(summary_model)["take_testosteroneTRUE", ]
  ci <- confint.default(model, "take_testosteroneTRUE", level = 0.95)

  # Append the results
  coefficients_df <- rbind(coefficients_df, data.frame(variable = var, 
                                                       coefficient = coef["Estimate"],
                                                       ci_lower = ci[1],
                                                       ci_upper = ci[2],
                                                       p_value = coef["Pr(>|t|)"]))
}
# Define significance levels for coloring
coefficients_df$significance <- ifelse(coefficients_df$p_value < 0.001, "p < 0.001",
                               ifelse(coefficients_df$p_value < 0.01, "p < 0.01",
                               ifelse(coefficients_df$p_value < 0.05, "p < 0.05", "p >= 0.05")))

# Plotting the coefficients as a forest plot
ggplot(coefficients_df, aes(x = reorder(variable,-coefficient), y = coefficient, color = significance)) +
  geom_point(size = 3) +  # Display the coefficient estimates as points
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +  # Display error bars for CIs
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_color_manual(values = c("p < 0.001" = "red", "p < 0.01" = "orange", "p < 0.05" = "blue", "p >= 0.05" = "grey")) +
  labs(title = "Effect of testosterone on Various Sleep and Activity Measures",
       subtitle = "Colored by Significance Levels of P-Value",
       x = "Coefficient of 'take_testosterone'",
       y = "Response Variable") +
  coord_flip()+
  theme_minimal()
```

```{r}
# Define the list of response variables
response_vars <- c("sleep_quality", "time_to_fall_asleep_numeric", "awake_minutes_numeric", 
                   "early_waking_numeric", "functional_impairment_numeric", "daytime_sleepiness_numeric")

# Initialize an empty data frame to store coefficients, confidence intervals, and p-values
coefficients_df <- data.frame(variable = character(), coefficient = numeric(), 
                              ci_lower = numeric(), ci_upper = numeric(), p_value = numeric())

# Loop through each response variable, fit the model, and extract the coefficient, CI, and p-value for 'take_testosterone'
for (var in response_vars) {
  data[[var]] <- (data[[var]] - min(data[[var]], na.rm = TRUE)) / (max(data[[var]], na.rm = TRUE) - min(data[[var]], na.rm = TRUE))
  formula <- as.formula(paste(var, "~ age_numeric + income_numeric + exercise_frequency_numeric + take_testosterone"))
  model <- lm(formula, data = data)
  
  # Extracting the coefficient, confidence interval, and p-value for 'take_testosterone'
  summary_model <- summary(model)
  coef <- coef(summary_model)["take_testosteroneTRUE", ]
  ci <- confint.default(model, "take_testosteroneTRUE", level = 0.95)

  # Append the results
  coefficients_df <- rbind(coefficients_df, data.frame(variable = var, 
                                                       coefficient = coef["Estimate"],
                                                       ci_lower = ci[1],
                                                       ci_upper = ci[2],
                                                       p_value = coef["Pr(>|t|)"]))
}
# Define significance levels for coloring
coefficients_df$significance <- ifelse(coefficients_df$p_value < 0.001, "p < 0.001",
                               ifelse(coefficients_df$p_value < 0.01, "p < 0.01",
                               ifelse(coefficients_df$p_value < 0.05, "p < 0.05", "p >= 0.05")))

# Plotting the coefficients as a forest plot
ggplot(coefficients_df, aes(x = reorder(variable,-coefficient), y = coefficient, color = significance)) +
  geom_point(size = 3) +  # Display the coefficient estimates as points
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +  # Display error bars for CIs
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_color_manual(values = c("p < 0.001" = "red", "p < 0.01" = "orange", "p < 0.05" = "blue", "p >= 0.05" = "grey")) +
  labs(title = "Effect of testosterone on Various Sleep and Activity Measures",
       subtitle = "Colored by Significance Levels of P-Value",
       x = "Coefficient of 'take_HRT'",
       y = "Response Variable") +
  coord_flip()+
  theme_minimal()
```


```{r}
# Define the list of response variables
response_vars <- c("sleep_quality", "time_to_fall_asleep_numeric", "awake_minutes_numeric", 
                   "early_waking_numeric", "functional_impairment_numeric", "daytime_sleepiness_numeric")

# Initialize an empty data frame to store coefficients, confidence intervals, and p-values
coefficients_df <- data.frame(variable = character(), coefficient = numeric(), 
                              ci_lower = numeric(), ci_upper = numeric(), p_value = numeric())

# Loop through each response variable, fit the model, and extract the coefficient, CI, and p-value for 'take_testosterone'
for (var in response_vars) {
  formula <- as.formula(paste(var, "~ age_numeric  + exercise_frequency_numeric + take_testosterone"))
  data[,var] = as.factor(data[,var])
 model <- polr(formula, data = data, Hess=TRUE)
  # Extracting the coefficient, confidence interval, and p-value for 'take_testosterone'
  coef_info <- coef(summary(model))
  p <- pnorm(abs(coef_info[, "t value"]), lower.tail = FALSE) * 2

  (coef_info <- cbind(coef_info, "p value" = p))
  ci <- confint(model, "take_testosteroneTRUE", level = 0.95)

  # Append the results
  coefficients_df <- rbind(coefficients_df, data.frame(variable = var, 
                                                       coefficient = coef_info["take_testosteroneTRUE", "Value"],
                                                       ci_lower = ci[1],
                                                       ci_upper = ci[2],
                                                       p_value = coef_info["take_testosteroneTRUE", "p value"]))
}

# Define significance levels for coloring
coefficients_df$significance <- ifelse(coefficients_df$p_value < 0.001, "p < 0.001",
                               ifelse(coefficients_df$p_value < 0.01, "p < 0.01",
                               ifelse(coefficients_df$p_value < 0.05, "p < 0.05", "p >= 0.05")))
# Plotting the coefficients as a forest plot
ggplot(coefficients_df, aes(x = reorder(variable, -coefficient), y = coefficient, color = significance)) +
  geom_point(size = 3) +  # Display the coefficient estimates as points
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +  # Display error bars for CIs
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_color_manual(values = c("p < 0.001" = "red", "p < 0.01" = "orange", "p < 0.05" = "blue", "p >= 0.05" = "grey")) +
  labs(title = "Effect of Testosterone on Various Sleep and Activity Measures",
       subtitle = "Colored by Significance Levels of P-Value",
       x = "Coefficient of 'take_testosterone'",
       y = "Response Variable") +
  coord_flip() +
  theme_minimal()
```
```{r}
# Define the list of response variables
response_vars <- c("sleep_quality", "time_to_fall_asleep_numeric", "awake_minutes_numeric", 
                   "early_waking_numeric", "functional_impairment_numeric", "daytime_sleepiness_numeric")

# Initialize an empty data frame to store coefficients, confidence intervals, and p-values
coefficients_df <- data.frame(variable = character(), coefficient = numeric(), 
                              ci_lower = numeric(), ci_upper = numeric(), p_value = numeric())

# Loop through each response variable, fit the model, and extract the coefficient, CI, and p-value for 'take_HRT'
for (var in response_vars) {
  formula <- as.formula(paste(var, "~ age_numeric  + exercise_frequency_numeric + take_HRT"))
  data[,var] = as.factor(data[,var])
 model <- polr(formula, data = data, Hess=TRUE)
  # Extracting the coefficient, confidence interval, and p-value for 'take_testosterone'
  coef_info <- coef(summary(model))
  p <- pnorm(abs(coef_info[, "t value"]), lower.tail = FALSE) * 2

  (coef_info <- cbind(coef_info, "p value" = p))
  ci <- confint(model, "take_HRTTRUE", level = 0.95)

  # Append the results
  coefficients_df <- rbind(coefficients_df, data.frame(variable = var, 
                                                       coefficient = coef_info["take_HRTTRUE", "Value"],
                                                       ci_lower = ci[1],
                                                       ci_upper = ci[2],
                                                       p_value = coef_info["take_HRTTRUE", "p value"]))
}

# Define significance levels for coloring
coefficients_df$significance <- ifelse(coefficients_df$p_value < 0.001, "p < 0.001",
                               ifelse(coefficients_df$p_value < 0.01, "p < 0.01",
                               ifelse(coefficients_df$p_value < 0.05, "p < 0.05", "p >= 0.05")))
# Plotting the coefficients as a forest plot
ggplot(coefficients_df, aes(x = reorder(variable, -coefficient), y = coefficient, color = significance)) +
  geom_point(size = 3) +  # Display the coefficient estimates as points
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +  # Display error bars for CIs
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_color_manual(values = c("p < 0.001" = "red", "p < 0.01" = "orange", "p < 0.05" = "blue", "p >= 0.05" = "grey")) +
  labs(title = "Effect of HRT on Various Sleep and Activity Measures",
       subtitle = "Colored by Significance Levels of P-Value",
       x = "Coefficient of 'take_HRTTRUE'",
       y = "Response Variable") +
  coord_flip() +
  theme_minimal()
```

```{r}
mod = glm(Physical_symptoms__e_g__pain__hot_flushes__sweats__racing_heart__needing_to_use_the_bathroom ~ age_numeric+income_numeric+take_testosterone, data=data, family = "binomial")
summary(mod)
```

```{r}
mod = glm(Physical_symptoms__e_g__pain__hot_flushes__sweats__racing_heart__needing_to_use_the_bathroom ~ age_numeric+income_numeric+take_HRT, data=data, family = "binomial")
summary(mod)
```